### 2020-10-23 latex/vlp.tex
Second edit of the VLP proof, much better now.
Time to move on to editing what Manfredi sent me on Oct 20

Before continuing, we must rigorously define what it means for a VNHC to inject
energy into the VLP.
We want the the VLP to behave in
such a way that momentum increases over time;
this intuition is captured mathematically by the following definitions.

\begin{defn}\label{defn:energy-injection}
   Let \(\dot{x} = f(x)\) be a planar ODE defined on \(D \subset \R^2\), with
   \(f(0) = (0,0)\).
   This system \textit{gains energy on \(D\)} if 
   \begin{enumerate}
      \item The origin is a repeller 
         (or equivalently, the origin is asymptotically stable in negative time).
      \item For all compact sets \(K \subset D\) and for almost every initial
         condition \(x(0) \in K\), there exists \(T > 0\) such
         that \(x(t) \not \in K \, (\forall t > T)\).
   \end{enumerate}
\end{defn}

\begin{defn}
   A regular VNHC \(h(\mathbf{q},\mathbf{p}) = 0\) 
   \textit{injects energy into the VLP} if the constrained dynamics gain energy
   everywhere on the constraint manifold, except possibly on a set of measure
   zero.
\end{defn}

@@

Supposing the initial condition of the VLP is not an equilibrium, we will show that the
constrained dynamics trace out a curve on the \((q,p)\)-plane which is diverging
from the origin.  
This implies that the momentum \(p\) is increasing in magnitude whenever the
curve hits the \(p\)-axis, which in turn means the VLP is gaining energy on
average.

\begin{thm}\label{thm:vlp-energy-stabilization}
   Define \(\theta := \arctan_2(p,q)\).
   Let \(L : \mathbb{S}^1 \rightarrow [\underline{l},\overline{l}]\) be a
   \(C^2\) function of \(\theta\).
   A regular VNHC of the form 
   \(h(\mathbf{q},\mathbf{p}) = l - L(\theta)\) injects energy into the VLP
   if there exists 
   \(l_\text{avg} \in [\underline{l},\overline{l}]\) such that 
   \begin{equation}\label{eqn:vlp-energy-gain-condition}
      \left(l_\text{avg} - L(\theta)\right)sin(2\theta) \geq 0 
      \text{ }\forall \theta \in \mathbb{S}^1
      ,
   \end{equation}
   with the property that the inequality is strict except at the
   coordinate axes.
   If the inequality is flipped, the VNHC dissipates energy. 
\end{thm}
\begin{proof}
   Let \(\Gamma = \mathbb{S}^1 \times \R \) be the constraint manifold for \(h\).
   In negative-time, the dynamics on \(\Gamma\) are
   \begin{equation}\label{eqn:negative-time-vlp}
      f(q,p) = 
      -\begin{bmatrix}
         \dot{q} \\ \dot{p}
      \end{bmatrix} = \begin{bmatrix}
         -\frac{p}{mL(\theta)^2} \\
         mgL(\theta)\sin(q)
      \end{bmatrix}
      .
   \end{equation}
   We begin by analyzing the upright equilibria \((\pm \pi, 0)\). 
   The linearization of (\ref{eqn:negative-time-vlp}) at these
   equilibria is given by
   \[
      df_{(\pm \pi,0)} = \begin{bmatrix}
         0 & -\frac{1}{mL(\pi)^2} \\
         -m g L(\pi) & 0 \\
      \end{bmatrix}
      ,
   \]
   which has eigenvalues at \(\lambda = \pm \sqrt{\frac{g}{L(\pi)}}\).
   The positive-time system has the same eigenvalues at these upright
   equilibria.
   We label the set of initial conditions converging to \((\pm \pi, 0)\) in
   positive time by
   \[
      \Pi_0^+ := \left\{ (q_0,p_0) \in \Gamma \mid \lim_{t \to \infty} (q(t),p(t)) = (\pm \pi, 0) \right\}
      ,
   \]
   and the set of initial conditions converging to \((\pm \pi,0)\)
   in negative time by \(\Pi_0^-\).
   Since only one eigenvalue is negative, the stable manifold theorem tells us
   that both \(\Pi_0^+\) and \(\Pi_0^-\) are one-dimensional \textbf{SOURCE?},
   which means they have measure zero within \(\Gamma\).

   To prove the origin of the positive-time system is a repeller, we will show
   the origin of the negative-time system is almost globally
   asymptotically stable.

   Choose, as a candidate Lyapunov function, the energy for the
   average-length pendulum with zero-potential at the bottom of the swing:
   \[
       E_\text{avg}(q,p) := \frac{1}{2}\frac{p^2}{m l_\text{avg}^2} 
           + m g l_\text{avg} (1-\cos(q))
      .
   \]
   This is positive definite at \((0,0)\) and has compact sublevel
   sets on \(\Gamma\). 
   The derivative of \(E_\text{avg}\) along (\ref{eqn:negative-time-vlp}) is
   \[
      \dot{E}_\text{avg} = \frac{g\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)}
              {l_\text{avg}^2L(\theta)^2}
      .
   \]
   The sign of \(\sin(q)p\) depends only on the quadrant in the
   \((q,p)\)-plane. In fact, it is easy to show that
   \[
      \sign{\sin(q)p} = \sign{\sin(2\theta)}
      .
   \]
   Furthermore, by Lemma \ref{lemma:sign-of-cube} we have
   \[ 
       \sign{L(\theta)^3 - l_\text{avg}^3} = \sign{L(\theta) - l_\text{avg}}
       .
   \]
   The derivative of \(E_\text{avg}\) is never positive since
   \begin{align}\label{eqn:vlp-proof-neg-inv}
    \sign{\dot{E}_\text{avg}} &= \sign{\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)} \\
                &= \sign{\sin(2\theta) \left(L(\theta) - l_\text{avg}\right)} \\
                &= -\sign{(l_\text{avg} - L(\theta))\sin(2\theta)} \\
                &\leq 0 \text{ (by assumption)}
      .
   \end{align}
   Thus, \(E_\text{avg}\) is a Lyapunov function, which means the negative-time system is
   stable \cite{lyapunov}. 
   Define \(D := \Gamma \backslash \Pi_0^-\) to be the domain of interest.
   Using a Krasovskii-LaSalle argument \cite{krasovskii_lasalle}, 
   we will prove that (\ref{eqn:negative-time-vlp}) asymptotically converges to
   \((0,0)\) if it is initialized on \(D\).
   First, we define the zero-set
   \[
      Z = \left\{(q,p) \in \Gamma \mid 
      \dot{E}_\text{avg}(q,p) = 0 \right\}
      .
   \]
   It is easy to show that \(Z\) is the union of the coordinate axes
   \[
      Z= \mathbb{S}^1 \times \left\{0\right\} \cup 
         \left\{0\right\} \times \R
      .
   \]
   Krasovskii-LaSalle tells us that solutions to (\ref{eqn:negative-time-vlp})
   converge to the largest invariant set in \(Z \cap D\).
   Since \(f(q,p) = (0,0)\) only when \((q,p) = (0,0)\),
   the origin is asymptotically stable on \(D\) for the negative-time system.
   Hence, the positive-time system is diverging away from the origin 
   in the \((q,p)\) plane, which means the origin is a repeller.

   We now prove the system escapes compact sets in finite time.
   Let \(x(t) = (q(t),p(t))\) be the orbit for the positive-time system.
   Let \(K \subset \Gamma\) be compact and assume \(x(0) \in K\backslash \Pi_0^+\).
   For a proof by contradiction, suppose that for all \(T > 0\) there exists \(t > T\)
   where \(x(t) \in K\).
   Let 
   \[
      k := \max\limits_{(q,p) \in K} E_\text{avg}(q,p)
      ,
   \] 
   be the largest energy value attained in \(K\), and let
   \[
      E_k := \left\{ (q,p) \in \Gamma \mid E_\text{avg}(q,p) \leq k \right\}
      ,
   \]
   be the energy sublevel set with that largest energy. 
   Note that \(E_k\) is compact, contains \(K\), and is negatively invariant (it
   is positively invariant for the negative-time system).
   Because \(K \subset E_k\), it follows that \(x(t) \in E_k\). 
   Then, since \(E_k\) is negatively invariant, the half orbit
   \(x([0,t])\) is contained fully within \(E_k\).
   In particular, \(x(T) \in E_k\); 
   but \(T\) was chosen arbitrarily, which implies \(x(t) \in E_k\) for all
   \(t \in [0,\infty[\).
   Since \(E_k\) is compact, \(x(t)\) has a positive limit set in
   \(E_k\) which (by Poincar\'{e}-Bendixson) is either an equilibrium or a
   closed orbit \textbf{SOURCE?}.
   The system cannot have closed orbits, since this would violate asymptotic
   stability of \((0,0)\) in negative time. 
   Hence, the positive limit set must be one of the equilibria. 
   The origin is a repeller, which means \(x(t)\) must be converging to 
   \((\pm \pi,0)\).
   This is a contradiction, since we assumed \(x(0) \not \in \Pi_0^+\).

   We have shown that orbits of the constrained dynamics will leave compact sets in
   finite time, including all sublevel sets of \(E_\text{avg}\). 
   The orbit \((q(t),p(t))\) is spiraling away from the origin and (for almost
   all initial conditions) will not hit the upper equilibrium.
   Thus, the VLP is gaining energy on average because its momentum is increasing.
   Flipping the inequality of (\ref{eqn:vlp-energy-gain-condition}) 
   means \(E_\text{avg}\) is a Lyapunov function for the positive time system;
   by the Krasovskii-LaSalle arguments presented earlier, the VLP loses energy
   while converging asymptotically to \((0,0)\).
 \end{proof}

>>>>

Before proving the VLP will gain energy under this VNHC, 
we must rigorously define what that means.
If any initial condition of the constrained dynamics converges to the point
\((q,p) = (0,0)\), the momentum must be decreasing over time, so the VLP will
clearly be losing energy.
A notion of energy injection must therefore require that the origin repels all
solutions.
We need to avoid closed orbits or limit cycles in the \((q,p)\)-plane,
so that the system can always reach higher speeds. 
Finally, we want the VLP's momentum to increase on average, regardless of
its initial value.
This intuition is captured mathematically by the following definition.

\begin{defn}\label{defn:energy-injection}
   Let \(D \subset \R^2\) with \(0 \in D\).
   Let \(f : D \rightarrow \R^2\) be locally Lipschitz with \(f(0) = 0\).
   The system described by \(\dot{x} = f(x)\) 
   \textit{gains energy on \(D\)} if 
   \begin{enumerate}
      \item The origin is a repeller 
         (or equivalently, the origin is asymptotically stable in negative time).
      \item For all compact sets \(K \subset D\) and for almost every initial
         condition \(x(0) \in K\), there exists \(T > 0\) such
         that \(x(t) \not \in K \, (\forall t > T)\).
   \end{enumerate}
   The system \textit{loses energy on \(D\)} if it gains energy in
   negative-time.
\end{defn}

The second property of Definition \ref{defn:energy-injection} allows the system
to have unstable equilibria, but not limit cycles nor closed orbits.
The next definition ties this notion of energy gain to VNHCs.

\begin{defn}
   A regular VNHC \(h(\mathbf{q},\mathbf{p}) = 0\) 
   \textit{injects (dissipates) energy} if the constrained dynamics gain (lose)
   energy everywhere on the constraint manifold, except possibly on a set of
   measure zero.
\end{defn}

@@

We will show that the constrained dynamics trace out a curve on the
\((q,p)\)-plane which is diverging from the origin.  
This implies that the momentum \(p\) is increasing in magnitude whenever the
curve hits the \(p\)-axis, which in turn means the VLP is gaining energy on
average.

\begin{thm}\label{thm:vlp-energy-stabilization}
   Define \(\theta := \arctan_2(p,q)\).
   Let \(L : \mathbb{S}^1 \rightarrow [\underline{l},\overline{l}]\) be a
   \(C^2\) function of \(\theta\).
   A regular VNHC of the form 
   \(h(\mathbf{q},\mathbf{p}) = l - L(\theta)\) injects energy into the VLP
   if there exists 
   \(l_\text{avg} \in [\underline{l},\overline{l}]\) such that 
   \begin{equation}\label{eqn:vlp-energy-gain-condition}
      \left(l_\text{avg} - L(\theta)\right)sin(2\theta) \geq 0 
      \text{ }\forall \theta \in \mathbb{S}^1
      ,
   \end{equation}
   with the property that the inequality is strict except at the
   coordinate axes.
   If the inequality is flipped, the VNHC dissipates energy. 
\end{thm}
\begin{proof}
   The constraint manifold for \(h\) is the cylinder \(\Gamma = \mathbb{S}^1 \times \R \). 
   In negative-time, the dynamics on \(\Gamma\) are
   \begin{equation}\label{eqn:negative-time-vlp}
      f(q,p) = 
      -\begin{bmatrix}
         \dot{q} \\ \dot{p}
      \end{bmatrix} = \begin{bmatrix}
         -\frac{p}{mL(\theta)^2} \\
         mgL(\theta)\sin(q)
      \end{bmatrix}
      .
   \end{equation}
   This has equilibria at \((q,p) = (0,0)\) and \((q,p) = (\pi,0)\) (the
   third ``equilibrium" \((-\pi,0)\) is the same as \((\pi,0)\) because 
   \(q \in \mathbb{S}^1\)).
   To prove the origin of the positive-time system (\ref{eqn:vlp-hamiltonian})
   is a repeller, we will show the origin of the negative-time system 
   (\ref{eqn:negative-time-vlp}) is almost everywhere globally asymptotically
   stable.

   Choose, as a candidate Lyapunov function, the energy for the
   average-length pendulum with zero-potential at the bottom of the swing:
   \[
       E_\text{avg}(q,p) := \frac{1}{2}\frac{p^2}{m l_\text{avg}^2} 
           + m g l_\text{avg} (1-\cos(q))
      .
   \]
   This is positive definite at \((0,0)\) and has compact sublevel
   sets on \(\Gamma\). 
   The derivative of \(E_\text{avg}\) along (\ref{eqn:negative-time-vlp}) is
   \[
      \dot{E}_\text{avg} = \frac{g\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)}
              {l_\text{avg}^2L(\theta)^2}
      .
   \]
   The sign of \(\sin(q)p\) depends only on the quadrant in the
   \((q,p)\)-plane. In fact, it is easy to show that
   \[
      \sign{\sin(q)p} = \sign{\sin(2\theta)}
      .
   \]
   Furthermore, by Lemma \ref{lemma:sign-of-cube} we have
   \[ 
       \sign{L(\theta)^3 - l_\text{avg}^3} = \sign{L(\theta) - l_\text{avg}}
       .
   \]
   The derivative of \(E_\text{avg}\) is never positive since
   \begin{align}\label{eqn:vlp-proof-neg-inv}
    \sign{\dot{E}_\text{avg}} &= \sign{\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)}
          \nonumber \\
          &= \sign{\sin(2\theta) \left(L(\theta) - l_\text{avg}\right)}
          \nonumber \\
          &= -\sign{(l_\text{avg} - L(\theta))\sin(2\theta)} 
          \nonumber \\
          &\leq 0 \text{ (by assumption)}
      .
   \end{align}
   Thus, \(E_\text{avg}\) is a Lyapunov function, which means the negative-time system is
   stable \cite{lyapunov}.

   The linearization of (\ref{eqn:negative-time-vlp}) at the upright equilibrium
   \((\pi, 0)\) is given by
   \[
      df_{(\pi,0)} = \begin{bmatrix}
         0 & -\frac{1}{mL(0)^2} \\
         -m g L(0) & 0 \\
      \end{bmatrix}
      ,
   \]
   which has eigenvalues at \(\lambda = \pm \sqrt{\frac{g}{L(0)}}\).
   The positive-time system (\ref{eqn:vlp-hamiltonian}) has the same eigenvalues
   at this equilibrium.
   Label the set of initial conditions converging to \((\pi, 0)\) in
   positive time by
   \[
      \Pi_0^+ := \left\{ (q_0,p_0) \in \Gamma \mid \lim_{t \to \infty} (q(t),p(t)) = (\pi, 0) \right\}
      ,
   \]
   and the set of initial conditions converging to \((\pi,0)\)
   in negative time by \(\Pi_0^-\).
   Since only one eigenvalue is negative, the stable manifold theorem tells us
   that both \(\Pi_0^+\) and \(\Pi_0^-\) are one-dimensional.
   While this implies the origin is not globally asymptotically stable for
   (\ref{eqn:negative-time-vlp}), we can still show it is asymptotically stable
   for almost every initial condition because \(\Pi_0^+\) and \(\Pi_0^-\) are of
   measure zero within \(\Gamma\).

   Define \(D := \Gamma \backslash \Pi_0^-\) to be the domain of interest.
   Using a Krasovskii-LaSalle argument \cite{krasovskii_lasalle}, 
   we will prove that (\ref{eqn:negative-time-vlp}) asymptotically converges to
   \((0,0)\) if it is initialized on \(D\).
   First, we define the zero-set
   \[
      Z = \left\{(q,p) \in \Gamma \mid 
      \dot{E}_\text{avg}(q,p) = 0 \right\}
      .
   \]
   It is easy to show that \(Z\) is the union of the coordinate axes
   \[
      Z= \mathbb{S}^1 \times \left\{0\right\} \cup 
         \left\{0\right\} \times \R
      .
   \]
   Krasovskii-LaSalle tells us that solutions to (\ref{eqn:negative-time-vlp})
   converge to the largest invariant set in \(Z \cap D\).
   Since \(f(q,p) = (0,0)\) on \(D\) only when \((q,p) = (0,0)\),
   the origin is asymptotically stable on \(D\) for the negative-time system.
   Hence, the positive-time system is diverging away from the origin 
   in the \((q,p)\) plane, which means the origin is a repeller.

   We now prove the system escapes compact sets in finite time.
   Let \(x(t) = (q(t),p(t))\) be the orbit for the positive-time system.
   Let \(K \subset \Gamma\) be compact and assume \(x(0) \in K\backslash \Pi_0^+\).
   For a proof by contradiction, suppose that for all \(T > 0\) there exists
   \(t' > T\) where \(x(t') \in K\).
   Let 
   \[
      k := \max\limits_{(q,p) \in K} E_\text{avg}(q,p)
      ,
   \] 
   be the largest energy value attained in \(K\), and let its sublevel set be
   \[
      E_k := \left\{ (q,p) \in \Gamma \mid E_\text{avg}(q,p) \leq k \right\}
      .
   \]
   All sublevel sets of \(E_\text{avg}\) are compact, negatively
   invariant (by (\ref{eqn:vlp-proof-neg-inv})), and \(E_k\) contains \(K\).
   We assumed \(x(t') \in K\), so it follows that \(x(t') \in E_k\). 
   Then, since \(E_k\) is negatively invariant, the half orbit
   \(x([0,t'])\) is contained fully within \(E_k\).
   In particular, \(x(T) \in E_k\);
   but \(T\) was chosen arbitrarily, which implies \(x(t) \in E_k\) for all
   \(t \in [0,\infty[\).
   Since \(E_k\) is compact, \(x(t)\) has a positive limit set in
   \(E_k\) which (by Poincar\'{e}-Bendixson) is either an equilibrium or a
   closed orbit \textbf{SOURCE?}.
   The system cannot have closed orbits, since this would violate asymptotic
   stability of \((0,0)\) in negative time. 
   Hence, the positive limit set must be one of the equilibria. 
   The origin is a repeller, which means \(x(t)\) must be converging to 
   \((\pi,0)\).
   This is a contradiction, since we assumed \(x(0) \not \in \Pi_0^+\).

   We have shown that orbits of the constrained dynamics will leave compact sets in
   finite time, including all sublevel sets of \(E_\text{avg}\). 
   The orbit \((q(t),p(t))\) is spiraling away from the origin and, for almost
   all initial conditions, will not converge to the upper equilibrium.
   Thus, the VLP is gaining energy on average because its momentum is increasing.
   Flipping the inequality of (\ref{eqn:vlp-energy-gain-condition}) 
   means \(E_\text{avg}\) is a Lyapunov function for the positive time system;
   by the same arguments presented here, the VLP loses energy while
   converging asymptotically to \((0,0)\).
 \end{proof}
